Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-01-29 paper/zhang-cook-laa-langrene-menendez.Rmd
writing on toy examples are now done

## Understanding the optimiser's coverage of the search space

@@

Several annotations have been made to help understanding this plot. The original space of the bases is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, PCA is performed to get the first two PC coordinates of all the bases. The search space in the 2D space is a circle with the origin being the PC coordinates of the zero matrix, and radius estimated as the largest distance between the origin and all the bases. The theoretical best basis is known with simulated data and can be labelled for comparing the how close the ending basis to the theoretical one for different optimisers. Various aesthetics, i.e. size, alpha and color, are applicable to each component to adjust for the presentation.

@@

```{r toy-pca-animated, fig.cap = "A selected number of frames from the animated PCA plots. With animation, it is easier to track the progression of each optimiser from the start to finish. A full video of this aimation can be found at \\url{https://vimeo.com/user132007430/review/504242845/b73f37175a}"}
```

Animated plots can be informative in diagnostics, especially in the case of PCA plot when the starting and ending of the search is not clear. Figure \ref{fig:toy-pca-animated} shows six frames of an animated version of Figure \ref{fig:toy-pca} and this time, it shows that `search_better` finds the optimum quicker than `search_geodesic`.

@@

```{r toy-tour, fig.cap="A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The tour animation allows for a more holistic view of the full space in high dimensions from different angles. A full video of this aimation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}"}
```

While viewing the bases on the reduced space via PCA shed some lights on the space the optimisers have explored, the visualisation on the original $p \times d$ dimension enables a more holistic stereoscopic view of the search. To view a high dimensional ($d \ge 3$) object on a screen, an approach is to play the rotation of the object in animation and this can be done via a regular grand tour. Compared to the PCA plot, the animated rotation (tour) displayed in Figure \ref{fig:toy-tour} gives a more well-rounded view of the search and one can view the curved region of the tour path from different angles, which may not be presented in the PCA plot. Also the grand tour animation encompasses the PCA projection since the rotation from PCA is just one angle that maximises the variance of the bases and the grand tour produces a sequence of angles that view the search from different directions. As an evidence, the last frame in Figure \ref{fig:toy-tour} is a frame select from the tour animation that is close to the PCA angle and the projection looks similar to the one in Figure \ref{fig:toy-pca}.

>>>>

## Understanding the optimiser's coverage of the search space {#toy-pca}

@@

Several annotations have been made to help understanding this plot. The original space of the bases is a high dimensional sphere and random bases on the sphere can be generated via the CRAN package \CRANpkg{geozoo} [@geozoo]. Along with the bases recorded during the optimisation and a zero basis, PCA is performed to get the first two PC coordinates of all the bases. The search space in the 2D space is a circle with the origin being the PC coordinates of the zero matrix, and radius estimated as the largest distance between the origin and all the bases. The theoretical best basis is known with simulated data and can be labelled for comparing the how close the ending basis to the theoretical one for different optimisers. Various aesthetics, i.e. size, alpha and color, are applicable to emphasize the crucial elements and adjust for the presentation. For example, anchor points and search points are less important and a smaller size and alpha is used; The interpolation paths are more informative if alpha increases from start to finish (that is, the path is more opaque towards the end).

@@

```{r toy-pca-animated, fig.cap = "Six selected frames from the animated PCA plots in the previous figure. With animation, the start and end position of each optimisation is easier to identify.  A full video of this aimation can be found at \\url{https://vimeo.com/user132007430/review/504242845/b73f37175a}"}
```

Animation is another display to show how the search porgresses from start to finish in the space and an  `animate = TRUE` argument is used to enable the animation. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that the SA search finds its end basis quicker than the PD search since SA has finished in the 5th frame while PD is still making more progression.

@@

```{r toy-tour, fig.cap="Six selected frames from the tour animation for viewing the same two optmisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \\url{https://vimeo.com/user132007430/review/504328122/9be84db563}"}
```

As mentioned in Section \ref{toy-pca}, the original $p \times d$ dimension space can be simulated via randomly generated bases via \CRANpkg{geozoo} [@geozoo] package. A tour plot rotates these random bases and the ones collected by the optimisation in the original space and hence provides a stereoscopic view of the search. While the PCA plot projects the bases from the direction that maximises the variance, the tour plot display of the original high dimensional space from various directions using animation. Figure \ref{fig:toy-tour} shows the tour plot of the same two optimisations in the 5D space.

### 2021-01-29 paper/zhang-cook-laa-langrene-menendez.tex
writing on toy examples are now done

\hypertarget{understanding-the-optimisers-coverage-of-the-search-space}{%
\subsection{Understanding the optimiser's coverage of the search
space}\label{understanding-the-optimisers-coverage-of-the-search-space}}

@@

Several annotations have been made to help understanding this plot. The
original space of the bases is a high dimensional sphere and random
bases on the sphere can be generated via the CRAN package
\CRANpkg{geozoo} \citep{geozoo}. Along with the bases recorded during
the optimisation and a zero basis, PCA is performed to get the first two
PC coordinates of all the bases. The search space in the 2D space is a
circle with the origin being the PC coordinates of the zero matrix, and
radius estimated as the largest distance between the origin and all the
bases. The theoretical best basis is known with simulated data and can
be labelled for comparing the how close the ending basis to the
theoretical one for different optimisers. Various aesthetics, i.e.~size,
alpha and color, are applicable to each component to adjust for the
presentation.

@@

\caption{A selected number of frames from the animated PCA plots. With animation, it is easier to track the progression of each optimiser from the start to finish. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

Animated plots can be informative in diagnostics, especially in the case
of PCA plot when the starting and ending of the search is not clear.
Figure \ref{fig:toy-pca-animated} shows six frames of an animated
version of Figure \ref{fig:toy-pca} and this time, it shows that
\texttt{search\_better} finds the optimum quicker than
\texttt{search\_geodesic}.

@@

\caption{A selected number of frames from the tour animation for viewing the 5D space of all the projection bases. The tour animation allows for a more holistic view of the full space in high dimensions from different angles. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

While viewing the bases on the reduced space via PCA shed some lights on
the space the optimisers have explored, the visualisation on the
original \(p \times d\) dimension enables a more holistic stereoscopic
view of the search. To view a high dimensional (\(d \ge 3\)) object on a
screen, an approach is to play the rotation of the object in animation
and this can be done via a regular grand tour. Compared to the PCA plot,
the animated rotation (tour) displayed in Figure \ref{fig:toy-tour}
gives a more well-rounded view of the search and one can view the curved
region of the tour path from different angles, which may not be
presented in the PCA plot. Also the grand tour animation encompasses the
PCA projection since the rotation from PCA is just one angle that
maximises the variance of the bases and the grand tour produces a
sequence of angles that view the search from different directions. As an
evidence, the last frame in Figure \ref{fig:toy-tour} is a frame select
from the tour animation that is close to the PCA angle and the
projection looks similar to the one in Figure \ref{fig:toy-pca}.

>>>>

\hypertarget{toy-pca}{%
\subsection{Understanding the optimiser's coverage of the search
space}\label{toy-pca}}

@@

Several annotations have been made to help understanding this plot. The
original space of the bases is a high dimensional sphere and random
bases on the sphere can be generated via the CRAN package
\CRANpkg{geozoo} \citep{geozoo}. Along with the bases recorded during
the optimisation and a zero basis, PCA is performed to get the first two
PC coordinates of all the bases. The search space in the 2D space is a
circle with the origin being the PC coordinates of the zero matrix, and
radius estimated as the largest distance between the origin and all the
bases. The theoretical best basis is known with simulated data and can
be labelled for comparing the how close the ending basis to the
theoretical one for different optimisers. Various aesthetics, i.e.~size,
alpha and color, are applicable to emphasize the crucial elements and
adjust for the presentation. For example, anchor points and search
points are less important and a smaller size and alpha is used; The
interpolation paths are more informative if alpha increases from start
to finish (that is, the path is more opaque towards the end).

@@

\caption{Six selected frames from the animated PCA plots in the previous figure. With animation, the start and end position of each optimisation is easier to identify.  A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504242845/b73f37175a}}\label{fig:toy-pca-animated}
\end{figure}
\end{Schunk}

Animation is another display to show how the search porgresses from
start to finish in the space and an \texttt{animate\ =\ TRUE} argument
is used to enable the animation. Figure \ref{fig:toy-pca-animated} shows
six frames from the animation of the PCA plot in Figure
\ref{fig:toy-pca}. An additional piece of information one can learn from
this animation is that the SA search finds its end basis quicker than
the PD search since SA has finished in the 5th frame while PD is still
making more progression.

@@

\caption{Six selected frames from the tour animation for viewing the same two optmisations as previous two figures. The tour animation allows for an appreciation of the search bases in the original high dimensional space. A full video of this aimation can be found at \url{https://vimeo.com/user132007430/review/504328122/9be84db563}}\label{fig:toy-tour}
\end{figure}
\end{Schunk}

As mentioned in Section \ref{toy-pca}, the original \(p \times d\)
dimension space can be simulated via randomly generated bases via
\CRANpkg{geozoo} \citep{geozoo} package. A tour plot rotates these
random bases and the ones collected by the optimisation in the original
space and hence provides a stereoscopic view of the search. While the
PCA plot projects the bases from the direction that maximises the
variance, the tour plot display of the original high dimensional space
from various directions using animation. Figure \ref{fig:toy-tour} shows
the tour plot of the same two optimisations in the 5D space.
